Clustering, profiling and plots

### 2024-09-03 data/02-access-and-prelim-plots.qmd
Draft WGiT Presentation and edits to manuscript

```{r reformat care-dest specific categories}
care_dest <- care_dest |> 
  mutate(Care_Category_Specific = case_when(
    Care_Category_Specific  == "Daycare_EarlyON" ~ "Daycare or EarlyON",
    Care_Category_Specific == "Community_Centre" ~ "Community or Recreation Centre",
    Care_Category_Specific == "Daycare_Centre" ~ "Daycare or EarlyON",
    Care_Category_Specific == "Recreation_Centre" ~ "Community or Recreation Centre",
    Care_Category_Specific == "School" ~ "School",
    Care_Category_Specific == "Long-Term_Care_Home" ~ "Long-Term Care or Retirement Home",
    Care_Category_Specific == "Retirement_Home" ~ "Long-Term Care or Retirement Home",
    Care_Category_Specific == "Senior_Centre" ~ "Senior Centre",
    Care_Category_Specific == "Seniors_Active_Living_Centre" ~ "Senior Centre",
    Care_Category_Specific == "Post_Office" ~ "Post Office",
    Care_Category_Specific == "Library" ~ "Library",
    Care_Category_Specific == "Banks" ~ "Bank or ATM",
    Care_Category_Specific == "Grocer_ConvenienceStore" ~ "Convenience Store",
    Care_Category_Specific == "Clinics" ~ "Hospital or Clinic",
    Care_Category_Specific == "Grocer_Retail" ~ "Grocer",
    Care_Category_Specific == "Dentist" ~ "Dentist",
    Care_Category_Specific == "Hosptial" ~ "Hospital or Clinic",
    Care_Category_Specific == "Hospital" ~ "Hospital or Clinic",
    Care_Category_Specific == "Pharmacy" ~ "Pharmacy",
    Care_Category_Specific == "Park" ~ "Park"),
  Care_Category = case_when(
    Care_Category == "Child_Centric" ~ "Dependent-centric",
    Care_Category == "Elder_Centric" ~ "Dependent-centric",
    Care_Category == "Errand_Centric" ~ "Errand-centric",
    Care_Category == "Grocery_Centric" ~ "Grocery-centric",
    Care_Category == "Health_Centric" ~ "Health-centric"))

@@

access_15 <- access_15 |> left_join(test_all, by=c("from_id", "Care_Category", "Care_Category_Specific")) |> ungroup()
rm(test_all)
access_15$V_isc |> sum(na.rm = TRUE) #sums to 2206 opportunities -- out of 2225 total opportunities. some opportunities canot be reached wtihin 15 min walk! 
```

@@

           Care_Category_Specific = rep(c("Park", "Daycare or EarlyON", "Convenience Store", "School",
                                          "Long-Term Care or Retirement Home", "Senior Centre",
                                          "Bank or ATM", "Library", 
                                          "Grocer", "Dentist", 
                                          "Pharmacy", "Post Office",
                                          "Community or Recreation Centre", "Hospital or Clinic"), length(R_PARCELS_CENTS_2020$ID)),
           copp_isc = rep(0, length(R_PARCELS_CENTS_2020$ID)*length(c("Park", "Daycare or EarlyON", "Convenience Store", "School",
                                          "Long-Term Care or Retirement Home", "Senior Centre",
                                          "Bank or ATM", "Library", 
                                          "Grocer", "Dentist", 
                                          "Pharmacy", "Post Office",
                                          "Community or Recreation Centre", "Hospital or Clinic"))),
           copp_isc_w = rep(0, length(R_PARCELS_CENTS_2020$ID)*length(c("Park", "Daycare or EarlyON", "Convenience Store", "School",
                                          "Long-Term Care or Retirement Home", "Senior Centre",
                                          "Bank or ATM", "Library", 
                                          "Grocer", "Dentist", 
                                          "Pharmacy", "Post Office",
                                          "Community or Recreation Centre", "Hospital or Clinic"))),

           V_isc = rep(0, length(R_PARCELS_CENTS_2020$ID)*length(c("Park", "Daycare or EarlyON", "Convenience Store", "School",
                                          "Long-Term Care or Retirement Home", "Senior Centre",
                                          "Bank or ATM", "Library", 
                                          "Grocer", "Dentist", 
                                          "Pharmacy", "Post Office",
                                          "Community or Recreation Centre", "Hospital or Clinic")))) |>
  mutate(from_id = as.character(from_id))

@@

#RECALL, categories are either 
# - "Dependent-centric" and include "Community or Recreation Centre", "Daycare or EarlyON", "Long-Term Care or Retirement Home", "Park", "School", "Senior Centre"
# - "Errand-centric" and include: "Bank or ATM", "Library", "Post Office"
# - "Grocery-centric" and include: "Convenience Store", "Grocer"
# - "Health-centric" and include: "Dentist", "Hospital or Clinic", "Pharmacy"
```
```{r}
access_15_byspecificcategory_DA |> group_by(Care_Category_Specific) |> summarise(V_isc = sum(V_isc))
care_dest |> st_drop_geometry() |> group_by(Care_Category_Specific) |> 
  summarise(count = n())

@@

```{r, eval=FALSE}
care_categories_plot <- 
 tm_shape(ham_bay, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(hydro_p_LakeOntario, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(Community_Boundaries, bbox=Community_Boundaries) + tm_polygons(col = "white", border.col = "black", lwd=1)+
  tm_shape(care_dest) + tm_dots("Care_Category_Specific", palette = RColorBrewer::brewer.pal(n = 4, name = 'Dark2'), size=0.025, title = "")+
  tm_facets("Care_Category", nrow = 2, free.scales= TRUE)+
  tm_scale_bar(position = c("left", "bottom"), breaks = c(0, 1, 5, 10, 20)) +
  tm_compass(position = c("left", "top"), size=0.9) +
  tm_layout(bg.color = "grey", panel.label.bg.color = "white", legend.position = c("right", "top"), legend.text.size = 0.5, legend.bg.color = "white", legend.bg.alpha = 0.4,
            panel.labels = c("1,265 Dependent-centric destinations", 
                             "158 Errand-centric destinations", 
                             "381 Grocery-centric destinations", 
                             "421 Health-centric destinations"),
            panel.label.size = 0.8)

>>>>

```{r reformat care-dest specific categories}
care_dest <- care_dest |> 
  mutate(Care_Category_Specific = case_when(
    Care_Category_Specific  == "Daycare_EarlyON" ~ "Daycare | EarlyON",
    Care_Category_Specific == "Community_Centre" ~ "Community | Rec. Centre",
    Care_Category_Specific == "Daycare_Centre" ~ "Daycare | EarlyON",
    Care_Category_Specific == "Recreation_Centre" ~ "Community | Rec. Centre",
    Care_Category_Specific == "School" ~ "School",
    Care_Category_Specific == "Long-Term_Care_Home" ~ "LTC | Ret. Home",
    Care_Category_Specific == "Retirement_Home" ~ "LTC | Ret. Home",
    Care_Category_Specific == "Senior_Centre" ~ "Senior Centre",
    Care_Category_Specific == "Seniors_Active_Living_Centre" ~ "Senior Centre",
    Care_Category_Specific == "Post_Office" ~ "Post Office",
    Care_Category_Specific == "Library" ~ "Library",
    Care_Category_Specific == "Banks" ~ "Bank or ATM",
    Care_Category_Specific == "Grocer_ConvenienceStore" ~ "Convenience Store",
    Care_Category_Specific == "Clinics" ~ "Hospital or Clinic",
    Care_Category_Specific == "Grocer_Retail" ~ "Grocer",
    Care_Category_Specific == "Dentist" ~ "Dentist",
    Care_Category_Specific == "Hosptial" ~ "Hospital or Clinic",
    Care_Category_Specific == "Hospital" ~ "Hospital or Clinic",
    Care_Category_Specific == "Pharmacy" ~ "Pharmacy",
    Care_Category_Specific == "Park" ~ "Park"),
  Care_Category = case_when(
    Care_Category == "Child_Centric" ~ "Dependent-centric",
    Care_Category == "Elder_Centric" ~ "Dependent-centric",
    Care_Category == "Errand_Centric" ~ "Errand-centric",
    Care_Category == "Grocery_Centric" ~ "Grocery-centric",
    Care_Category == "Health_Centric" ~ "Health-centric"))

@@

access_15 <- access_15 |> left_join(test_all, by=c("from_id", "Care_Category", "Care_Category_Specific")) |> ungroup()
rm(test_all)
access_15$V_isc |> sum(na.rm = TRUE) #sums to 2206 opportunities -- out of 2225 total opportunities. some opportunities cannot be reached within 15 min walk! 
```

@@

           Care_Category_Specific = rep(c("Park", "Daycare | EarlyON", "Convenience Store", "School",
                                          "LTC | Ret. Home", "Senior Centre",
                                          "Bank or ATM", "Library", 
                                          "Grocer", "Dentist", 
                                          "Pharmacy", "Post Office",
                                          "Community | Rec. Centre", "Hospital or Clinic"), length(R_PARCELS_CENTS_2020$ID)),
           copp_isc = rep(0, length(R_PARCELS_CENTS_2020$ID)*length(c("Park", "Daycare | EarlyON", "Convenience Store", "School",
                                          "LTC | Ret. Home", "Senior Centre",
                                          "Bank or ATM", "Library", 
                                          "Grocer", "Dentist", 
                                          "Pharmacy", "Post Office",
                                          "Community | Rec. Centre", "Hospital or Clinic"))),
           copp_isc_w = rep(0, length(R_PARCELS_CENTS_2020$ID)*length(c("Park", "Daycare | EarlyON", "Convenience Store", "School",
                                          "LTC | Ret. Home", "Senior Centre",
                                          "Bank or ATM", "Library", 
                                          "Grocer", "Dentist", 
                                          "Pharmacy", "Post Office",
                                          "Community | Rec. Centre", "Hospital or Clinic"))),

           V_isc = rep(0, length(R_PARCELS_CENTS_2020$ID)*length(c("Park", "Daycare | EarlyON", "Convenience Store", "School",
                                          "LTC | Ret. Home", "Senior Centre",
                                          "Bank or ATM", "Library", 
                                          "Grocer", "Dentist", 
                                          "Pharmacy", "Post Office",
                                          "Community | Rec. Centre", "Hospital or Clinic")))) |>
  mutate(from_id = as.character(from_id))

@@

#RECALL, categories are either 
# - "Dependent-centric" and include "Community | Rec. Centre", "Daycare | EarlyON", "LTC | Ret. Home", "Park", "School", "Senior Centre"
# - "Errand-centric" and include: "Bank or ATM", "Library", "Post Office"
# - "Grocery-centric" and include: "Convenience Store", "Grocer"
# - "Health-centric" and include: "Dentist", "Hospital or Clinic", "Pharmacy"
```
```{r}
access_15_byspecificcategory_DA |> group_by(Care_Category_Specific) |> summarise(V_isc = sum(V_isc))
care_dest |> st_drop_geometry() |> group_by(Care_Category_Specific) |> 
  summarise(count = n())

@@

```{r, eval=FALSE}
care_categories_plot <- 
 tm_shape(ham_bay, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(hydro_p_LakeOntario, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(Community_Boundaries, bbox=Community_Boundaries) + tm_polygons(col = "white", border.col = "black", lwd=1)+
  tm_shape(care_dest) + tm_dots("Care_Category_Specific", palette = RColorBrewer::brewer.pal(n = 4, name = 'Dark2'), size=0.025, title = "")+
  tm_facets("Care_Category", nrow = 2, free.scales= TRUE)+
  tm_scale_bar(position = c("left", "bottom"), breaks = c(0, 1, 5, 10, 20)) +
  tm_compass(position = c("left", "top"), size=0.9) +
  tm_layout(bg.color = "grey", panel.label.bg.color = "white", legend.position = c("right", "top"), legend.text.size = 1, legend.bg.color = "white", legend.bg.alpha = 0.4,
            panel.labels = c("1,265 Dependent-centric destinations", 
                             "158 Errand-centric destinations", 
                             "381 Grocery-centric destinations", 
                             "421 Health-centric destinations"),
            panel.label.size = 1.1)

### 2024-09-03 data/03-clustering-and-profiling.qmd
Draft WGiT Presentation and edits to manuscript

pop_descriptive_plots_1 <- tmap_arrange(parcels_DA_plot, pop_DA_plot, poprate_DA_plot, nrow=3)
tmap_save(pop_descriptive_plots_1,
          file= file.path(here("figures/pop_descriptive_plots_1.png")) , asp = 1/0.6,
          dpi=300)
```

@@

SClabel_copp_plot <- tm_shape(ham_bay, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(df_DA) +
  tm_polygons("max_label",
              border.alpha = 0.2,
              palette = c("#44ce1b","#bbdb44", "#f7e379", "#f2a134","#e51f1f"),
              labels = (c("Very caring - very complete (SC2)","Caring - Complete (SC1)",
              "Somewhat caring - complete (SC4)","Somewhat caring - somewhat complete (SC5)", 
              "Somewhat caring - somewhat complete (SC3)")),
              title = "Prevelant supercluster") +
  tm_shape(Community_Boundaries) + tm_polygons(alpha=0)+
  tm_scale_bar(position = c("left", "bottom"), breaks = c(0, 1, 5, 10, 15)) +
  tm_compass(position = c("left", "top"), size=1.0)+
  tm_layout(legend.bg.color = "white", legend.bg.alpha = 0.4,
            bg.color = "grey",
            legend.position = c("right","top"))
SClabel_copp_plot
SCLab_and_total_copp_plot <- tmap_arrange(SClabel_copp_plot, total_copp_plot, nrow=2)

tmap_save(SCLab_and_total_copp_plot, 
          file= file.path(here("figures/SCLab_and_total_copp_plot.png")),
          dpi=300)
```

@@

freq_sc
```

@@

#Train a decision tree using the data:
classdemo.som <- rpart(data = df,
                      formula = scluster ~ (Income_bottom + Age_0to14 + One_parent_families + Vis_minority + Not_in_labforce+ Income_gini+walk_towork+Owner_hld_CHN+Hld_ownership+Tenant_subsi_hld+Tenant_hld_CHH+LICO),
                      weights = Population
                      )
# # Visualize the decision tree:
# rpart.plot(classdemo.som, cex = 0.5, type = 0, varlen = 8)
```

@@

  # # pass list to gglist containing all ggplot components we want to plot for each
  # # (default: terminal) node
  geom_node_plot(gglist = list(geom_bar(aes(x = "", fill = scluster),
                                        position = position_fill()),
                               xlab(""),
                               ylab(""),
                               theme_void(),
                               theme(axis.text = element_blank())))
legd <- cowplot::get_legend(freq_sc)
```

```{r}
#plot and save
png(file= file.path(here("figures/access_sc_profiles.png")),width= 2404, height= 1600, units="px",  res=300)
plot_grid(party_plot, legd, nrow = 1, rel_widths = c(1, 0.2))
dev.off()
```

>>>>

pop_DA_plot_2 <-
 tm_shape(ham_bay, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(access_15_DA) +
  tm_polygons("Population",
              border.alpha = 0.2,
              palette = "Greens", style = "quantile", n=4,
              title = "Population",
              legend.is.portrait = TRUE) +
  tm_shape(Community_Boundaries) + tm_polygons(alpha=0)+
  tm_scale_bar(position = c("left", "bottom"), breaks = c(0, 1, 5, 10, 15)) +
  tm_compass(position = c("left", "top"))+
   tm_layout(bg.color = "grey", panel.label.bg.color = "white", legend.position = c("right", "top"), legend.bg.color = "white", legend.bg.alpha = 0.4)

pop_descriptive_plots_1 <- tmap_arrange(parcels_DA_plot, pop_DA_plot, poprate_DA_plot, nrow=3)
tmap_save(pop_descriptive_plots_1,
          file= file.path(here("figures/pop_descriptive_plots_1.png")) , asp = 1/0.6,
          dpi=300)

tmap_save(pop_DA_plot_2,
          file= file.path(here("figures/pop_descriptive_plot_onlypop.png")),
          dpi=300)
```

@@

SClabel_copp_plot <- tm_shape(ham_bay, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0) +
  tm_shape(hydro_p_LakeOntario, bbox=Community_Boundaries) + tm_polygons(col="skyblue", border.alpha = 0)+
  tm_shape(df_DA) +
  tm_polygons("max_label",
              border.alpha = 0.2,
              palette = c("#44ce1b","#bbdb44", "#f7e379", "#f2a134","#e51f1f"),
              labels = (c("Very caring - very complete",
                          "Caring - Complete",
              "Somewhat caring - complete",
              "Somewhat caring - somewhat complete", 
              "Somewhat caring - somewhat complete")),
              title = "Neighbourhood typology") +
  tm_shape(Community_Boundaries) + tm_polygons(alpha=0)+
  tm_scale_bar(position = c("left", "bottom"), breaks = c(0, 1, 5, 10, 15)) +
  tm_compass(position = c("left", "top"), size=1.0)+
  tm_layout(legend.bg.color = "white", legend.bg.alpha = 0.4,
            bg.color = "grey",
            legend.position = c("right","top"))
SClabel_copp_plot
SCLab_and_total_copp_plot <- tmap_arrange(SClabel_copp_plot, total_copp_plot, nrow=2)

tmap_save(SCLab_and_total_copp_plot, 
          file= file.path(here("figures/SCLab_and_total_copp_plot.png")),
          dpi=300)

tmap_save(SClabel_copp_plot, 
          file= file.path(here("figures/SClabel_copp_plot.png")),
          dpi=300)
```

@@

freq_sc

#with new names and colours
freq_sc_renamed <- ggplot(access_15_wide |> 
         mutate(scluster = factor(scluster))) +
  geom_bar(aes(x = "", 
               group = scluster, 
               fill = scluster)) +
  scale_fill_manual(name = "Neighbourhood typologies",
                    values = c("#44ce1b","#bbdb44", "#f7e379", "#f2a134","#e51f1f"),
                    labels = c("Very caring - very complete","Caring - complete","Somewhat caring - not complete","Somewhat caring - complete","Not caring - not complete"))
```

@@

#Train a decision tree using the data (renamed!):
classdemo.som <- rpart(data = df,
                      formula = scluster ~ (Income_bottom + Age_0to14 + One_parent_families + Vis_minority + Not_in_labforce+ Income_gini+walk_towork+Owner_hld_CHN+Hld_ownership+Tenant_subsi_hld+Tenant_hld_CHH+LICO),
                      weights = Population
                      )
# # Visualize the decision tree:
# rpart.plot(classdemo.som, cex = 0.5, type = 0, varlen = 8)
```

@@

  # # pass list to gglist containing all ggplot components we want to plot for each
  # # (default: terminal) node
  geom_node_plot(gglist = list(geom_bar(aes(x = "", fill = scluster),
                                        position = position_fill()),
                               #xlab(""),
                               #ylab(""),
                               theme_void(),
                               scale_fill_manual(values = c("#bbdb44","#44ce1b",  "#f2a134", "#f7e379","#e51f1f")),
                               theme(axis.text = element_blank())))

legd_renamed <- cowplot::get_legend(freq_sc_renamed)
party_plot
```

```{r, eval=FALSE}
#plot and save
# NOTE: edited this in paint.. be careful saving over
png(file= file.path(here("figures/access_sc_profiles_renamed.png")),width= 2600, height= 1600, units="px",  res=300)
plot_grid(party_plot, legd_renamed, nrow = 1, rel_widths = c(1, 0.3))
dev.off()
```
